---
title: "gdbR: R Client for the Genomics Platform Database"
author: "Valerie Obenchain"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gdbR: R Client for the Genomics Platform Database}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 4
---

#```{r biocstyle, echo=FALSE, warning=FALSE}
#suppressPackageStartupMessages({
#    library(GenomicRanges)
#    library(SummarizedExperiment)
#    library(RaggedExperiment)
#})
#```

# Introduction

The `gdbR` package provides an client interface to datasets in the 
Genomics Platform database.

The package provides functions to browse and download existing datasets
as well as save new datasets.

```{r load-libs}
library(gdbR)
```

# Explore dataset metadata

## General browsing

The `gdbDatasets()` function returns a list of datasets your user id has
permission to access.
```{r gdb-create}
gdb <- gdbDatasets()
```

This object displays the "DS" identifier and title. You can interact with it much as
you would a list. A '[' subset will return a new list with the specified elements
and the double bracket subset '[[' returns detailed metadata for an individual dataset.
```{r gdb-subset}
## The full list
gdb

## Subset the list
gdb[1:3]

## Extract metadata for a particular dataset
gdb[[5]]
```

A number of helpers are available to extract metadata terms from the
`GDBProjectList`. These terms can be used to subset the list or in 
a full database search. (NOTE: Full search not implemented yet.)
See the ?`GDBProjectList` man page for a full list of helper
functions.

* `availableDatasetIDs`: 
  A character vector of dataset ids assigned by the database.
  Dataset ids are unique.
* `availableTitles`: 
  A character vector of dataset titles. Titles are unique.
* `availableOrganisms`: 
  A character vector of unique organisms. Organisms are not
  unique across datasets. 
* `availablePlatforms`: 
  A character vector of unique platforms. Platforms are not
  unique across datasets.
* `availableSources`: 
  A character vector of unique data sources, e.g., ExpressionPlot, 
  FDB, GEO. Sources are not unique across datasets. 
* `availableSourceIDs`: 
  A chacter vector of unique source ids, e.g., PRJ001, NGS001, 
  FDB001. Source IDs are not unique across datsets.

```{r gdb-discovery}
head(availableDatasetIDs(gdb))
head(availableSourceIDs(gdb))
```

## ExpressionPlot Projects

All ExpressionPlot Projects were imported to the Genomics Platform 
database and assigned a unique dataset identifier (e.g., "DS" id). The
ExpressionPlot PRJ id is considered a "source id" and can be retrieved
with `availableEPProjects()`.

```{r gdb-availableEPProjects}
head(availableEPProjects(gdb))
```

Some functions in `gdbR` accept either a DS id or PRJ id as input but
the majority expect the DS id. A PRJ can be mapped to the associated
DS id with `mapEPtoDS`. The return object is a `data.frame` with 
ep_id, dataset_id and dataset_title.

```{r gdb-mapEPtoDS}
mapEPtoDS("PRJ0031967")
```

## FireDB and NGS Pipeline Projects

TBD

# Full database search

TBD

# Download existing datasets

Datasets can be downloaded as `MultiAssayExperiment` objects. Each dataset has
one or more assays which is associated with one or more resources. In terms of
`MultiAssayExperiment` structure, each assay becomes a `SummarizedExperiment` and
each associated resource is a matrix in the @assays slot of the `SummarizedExperiment`.

Retrieve the dataset with the DS identifier:

```{r gdb-getDatasetAsMAE-ds}
head(availableDatasetIDs(gdb))
mae <- getDatasetAsMAE("DS000003074")
```

Source identifiers can also be used for retrieval:
```{r gdb-getDatasetAsMAE-source}
head(availableSourceIDs(gdb))
mae <- getDatasetAsMAE("NGS2480")
```

This mae has a single `SummarizedExperiment` with two resources in
the assays slot.

```{r gdb-getDatasetAsMAE-explore}
names(mae)
se <- experiments(mae)[[1]]
assays(se)
```

# Save new datasets




# Versioning and Relationships

Datasets consist of assays, resources and samples.
The assay data are immutable but other characteristics can be
modified. When a change is made either a new version is created or an entirely
new dataset is created.

Actions that result in a new version of the dataset:
- Modifying the dataset metadata, e.g., changing the title, abstract
- Modifying the assay metadata, e.g., changing the title 
- Modifying the sample metadata, e.g., adding new covariates

Actions that result in a new dataset:
- Change in sample membership (samples are added or removed)
- Merging or subsetting datasets

 
Relationships: TBD

# Dataset access

## Read access

By default read access to a dataset is private.

* Private read: The dataset is only visable by people listed in 
`peopleAccess(GDBParam)`.
* Public read: The dataset is visable by everyone in the Bioinfo group
(others?). Group membership is checked by d-list/LDAP.

When adding a new dataset, read access can be set to public with 
`isPublic=TRUE` in the `GDBParam` object.

TBD: How to change read access in an existing dataset.

## Write access

Write access is restricted to persons listed in `peopleAccess(GDBParam)`.
The `peopleAccess` getter returns a data.frame of user id and roles.
Any user specified in this data.frame, reguardless of role, will have 
write access to the dataset.

Three roles are possible in the `peopleAccess` data.frame:: 
* Creator: The developer who adds the new dataset.
* Analyst: The bioinformatician analyzing the data
* PI: The lab head or person leading the experiment

# Shiny App

TBD

# REST API

## Import Endpoints

TBD

## Export Endpoints

http://gred-rstudio-p01.sc1.roche.com:3939/gdb

